Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-24 mr-review.Rmd
found citation for mode-based MR
added text about hartwig 2017 - mode based estimation

@smith_mendelian_2004 recognized that many genetic variants individually explain a 
small proportion of the variation in a trait, and, thus, sufficient statistical 
power would require sample sizes in the tens of thousands 
[@schatzkin2009mendelian;@burgess_mendelian_2013]. @palmer_using_2012 partially 
resolved this issue by proposing use of multiple SNP instruments in a single MR 
analysis. @pierce_power_2011 reasoned that a single causal estimate derived from 
a collection of causal SNPs would have greater precision than an estimate derived 
from only one SNP [@burgess_mendelian_2013]. At the same time, sharing of GWAS 
summary statistics was encouraged as a best practice [@nature2012asking]. 
@harbord_severity_2013 recognized that one could calculate the ratio estimate of 
the causal effect from GWAS summary statistics, and @burgess_mendelian_2013 devised 
a strategy for determining a single causal effect from summary statistics for a 
collection of SNPs. @burgess_mendelian_2013 termed their method "inverse variance 
weighting" because each causal SNP's effect is weighted by the inverse of the variance 
of the ratio estimator, and the overall causal effect is the sum of the weighted SNP
causal effects.

@@

@bowden_consistent_2016, noting that consistency with IVW MR is not robust to 
invalid instrument use, developed a method for consistent MR analysis with  
weighted median-based estimators. Building on research from @han2008detecting, 
@bowden_consistent_2016 devised a weighted median-based estimator. Its breakdown point is 50\%, 
meaning that up to 50\% of SNP instruments can be invalid while maintaining consistency in estimation 
[@hampel1971general]. Specifically, let $\tilde \beta_j$ denote the $j^{th}$ ordered ratio estimate 
statistic, from least to greatest. If an odd number, say $2K +1$, of SNP instruments is used, then 
the simple median estimator is defined as the $K + 1$ ordered ratio estimate. For an even number, say 
$2K$, number of SNP instruments, choose the midpoint between the two middle-ranking ordered ratio 
estimates, $\frac{\tilde \beta_K + \tilde \beta_{K + 1}}{2}$.

@@

weighted mode estimator

>>>>

@smith_mendelian_2004 recognized that many genetic variants individually explain a 
small proportion of the variation in a trait, and, thus, sufficient statistical 
power would require sample sizes in the tens of thousands 
[@schatzkin2009mendelian;@burgess_mendelian_2013]. @palmer_using_2012 partially 
resolved this issue by proposing use of multiple SNP instruments in a single MR 
analysis. @pierce_power_2011 reasoned that a single causal estimate derived from 
a collection of causal SNPs would have greater precision than an estimate derived 
from only one SNP [@burgess_mendelian_2013]. At the same time, sharing of GWAS 
summary statistics was encouraged as a best practice [@nature2012asking]. 
@harbord_severity_2013 recognized that one could calculate the ratio estimate of 
the causal effect from GWAS summary statistics, and @burgess_mendelian_2013 devised 
a strategy for determining a single causal effect from summary statistics for a 
collection of SNPs. @burgess_mendelian_2013 termed their method "inverse variance 
weighting" (IVW) because each causal SNP's effect is weighted by the inverse of the variance 
of the ratio estimator, and the overall causal effect is the sum of the weighted SNP
causal effects.

@@

@greco_m_detecting_2015

@bowden_mendelian_2015

@bowden_consistent_2016, noting that consistency with IVW MR is not robust to 
invalid instrument use, developed a method for consistent MR analysis with  
weighted median-based estimators, building on research from @han2008detecting. 
Its breakdown point is 50\%, 
meaning that up to 50\% of SNP instruments can be invalid while maintaining consistency in estimation 
[@hampel1971general]. Specifically, let $\tilde \beta_j$ denote the $j^{th}$ ordered ratio estimate 
statistic, from least to greatest. If an odd number, say $2K +1$, of SNP instruments is used, then 
the simple median estimator is defined as the $K + 1$ ordered ratio estimate. For an even number, say 
$2K$, number of SNP instruments, choose the midpoint between the two middle-ranking ordered ratio 
estimates, $\frac{\tilde \beta_K + \tilde \beta_{K + 1}}{2}$.

@@

@kang_instrumental_2016

@hartwig_robust_2017 developed a weighted mode-based MR causal effect estimation method for MR analysis with GWAS summary data. 
They first calculate Wald ratio estimates for every instrumental SNP. They then apply 
smoothing to the empirical distribution of ratio estimates. The mode of this smoothed 
distribution is the simple mode-based estimate of the causal effect 
[@hartwig_robust_2017]. The inverse variance-weighted mode-based estimate is obtained by 
weighting the empirical distribution of ratio estimates by the inverse variance of each 
estimate. @hartwig_robust_2017 compared their weighted and unweighted mode-based 
estimators with weighted and unweighted median-based estimators, MR Egger regression, and 
IVW. They found that their mode-based estimators demonstrated less bias and lower type I error rates than other estimators in some simulation settings. The mode-based estimators possessed lower statistical power to detect a causal effect compared to the IVW and weighted median-based methods, but their power exceeded that of MR Egger regression [@hartwig_robust_2017]. The mode-based estimators consistently estimated the causal effect when the mode across SNP instruments of the horizontal pleiotropy effects was zero [@hartwig_robust_2017]. In this manner, the mode-based estimators demonstrated a greater robustness to horizontal pleiotropy than did the other estimators. @hartwig_robust_2017
